---
title: "FLASH-seq RN26038 — data quality"
subtitle: "R / plotthis / scplotter rendering of `res/flashseq/`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    df_print: paged
    fig_width: 9
    fig_height: 3.6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dev = "png", dpi = 130, fig.align = "center")
```

## What this document is, and what it is not

This is the R rendering of the FLASH-seq QC. **It computes nothing.** Every number
below is read from the TSVs in `res/flashseq/`, which were written by the Python
scripts `00_collect_qc.py` … `06_trim_options.sh`. That is deliberate: two reports
that each recomputed their own numbers could disagree, and then neither would be
trustworthy. The only arithmetic done here is on the RSEM TPM matrix, for the
replicate scatters and the PCA — and the *r* values annotated on those scatters
still come from `replicate_concordance.tsv` rather than being recomputed.

**A note on `scplotter`.** The request was to visualise with `scplotter`, and the
figures below are drawn with **`plotthis`** — which *is* scplotter's plotting
engine, by the same author (Panwen Wang), so the theming, palettes and layout are
scplotter's. The reason is that `scplotter`'s own exported functions
(`CellDimPlot`, `CellStatPlot`, `FeatureStatPlot`, `Clonal*Plot`, `Spat*Plot`, …)
all take a **single-cell or spatial object** — Seurat, SingleCellExperiment or
scRepertoire — and every table in `res/flashseq/` is a *bulk, per-library QC
summary*: ten rows, one per library. There is no cell dimension for them to plot.

Section 9 is the exception: there, the ten libraries genuinely can be treated as
ten samples in a Seurat object, and `scplotter::CellDimPlot` is used for what it
is built for.

```{r libs}
suppressPackageStartupMessages({
  library(plotthis)
  library(ggplot2)
})

ROOT <- Sys.getenv("FS_ROOT", "/nemo/lab/turnerj/working/guangxin/vasaseq")
CODE <- file.path(ROOT, "code/flashseq")
RES  <- Sys.getenv("FS_OUT", file.path(ROOT, "res/flashseq"))
FIG  <- file.path(RES, "figures_R")
dir.create(FIG, showWarnings = FALSE, recursive = TRUE)

# Saved alongside each figure so the PDFs can be dropped into a slide deck, and
# so the R and Python renderings can be compared panel by panel.
save_fig <- function(p, name, width = 9, height = 3.6) {
  ggplot2::ggsave(file.path(FIG, paste0(name, ".pdf")), p,
                  width = width, height = height, device = grDevices::cairo_pdf)
  p
}

cat("plotthis  ", as.character(packageVersion("plotthis")), "\n")
cat("scplotter ", if (requireNamespace("scplotter", quietly = TRUE))
                   as.character(packageVersion("scplotter")) else "not installed", "\n")
cat("R         ", R.version.string, "\n")
```

## 0. The design — and why it has to be carried through everything

The run is **not** ten comparable samples. It is an **input-amount titration in
duplicate**, 30 ng down to 30 pg — a 1000-fold range. A mammalian cell carries
roughly 10–30 pg of total RNA, so **30 pg is the single-cell-equivalent rung**,
and it is the rung that matters for comparing against VASA-seq.

Nothing inside `data/flashseq/` records this. It comes from the RN26038 LIMS sheet
and lives in `code/flashseq/sample_metadata.tsv`, which every script joins against.
`qc_verdict` is the **decision** taken on 2026-07-27 to exclude A8.

```{r meta}
meta <- read.delim(file.path(CODE, "sample_metadata.tsv"), comment.char = "#")
# Everything in this document is ordered by input amount, high to low.
meta <- meta[order(-meta$input_pg), ]
ORDER  <- meta$library
SHORT  <- sub("ZHA8833", "", ORDER)
LEVELS <- unique(meta$input_amount)

# One short label per library, used as the x axis throughout.
shorten <- function(x) factor(sub("ZHA8833", "", x), levels = SHORT)
as_level <- function(x) factor(x, levels = LEVELS)

meta$short        <- shorten(meta$library)
meta$input_amount <- as_level(meta$input_amount)
EXCLUDED <- meta$library[meta$qc_verdict == "exclude"]

meta[, c("library", "input_amount", "replicate", "well", "qc_verdict", "qc_note")]
```

```{r palettes}
# One colour per input level, so a panel can be read by input amount alone.
PAL <- "Spectral"

# plotthis defaults to a near-square panel with the legend on top. With ten
# libraries on the x axis that runs the labels together into "A1A2A3...". These
# two overrides fix it, and live here rather than being repeated per figure so
# every per-library bar chart in the document stays consistent.
bar_lib <- function(...) BarPlot(..., legend.position = "right", aspect.ratio = 0.55)
```

## 1. Sequencing and alignment

From `qc_summary.tsv`, written by `00_collect_qc.py` out of the MultiQC tables.

```{r qc-table}
qc <- read.delim(file.path(RES, "qc_summary.tsv"))
qc <- qc[match(ORDER, qc$library), ]
qc$short        <- shorten(qc$library)
qc$input_amount <- as_level(qc$input_amount)

qc_show <- qc[, c("library", "input_amount", "replicate", "well", "qc_verdict",
                  "raw_pairs", "star_uniquely_mapped_pct", "star_multimapped_pct",
                  "star_unmapped_tooshort_pct", "rsem_alignable_pct",
                  "picard_dup_pct", "qualimap_intergenic_pct", "qualimap_5_3_bias",
                  "pct_MT", "insert_size_avg")]
qc_show$raw_pairs_M <- round(qc_show$raw_pairs / 1e6, 1)
qc_show$raw_pairs <- NULL
qc_show
```

```{r fig1, fig.width=11, fig.height=3.4}
panels <- c(star_uniquely_mapped_pct   = "STAR uniquely mapped (%)",
            star_unmapped_tooshort_pct = "STAR unmapped, too short (%)",
            picard_dup_pct             = "duplicate reads (%)")

fig1_dat <- do.call(rbind, lapply(names(panels), function(col)
  data.frame(short = qc$short, input_amount = qc$input_amount,
             metric = panels[[col]], value = qc[[col]])))
fig1_dat$metric <- factor(fig1_dat$metric, levels = unname(panels))

p <- bar_lib(fig1_dat, x = "short", y = "value", fill_by = "input_amount",
             facet_by = "metric", facet_scales = "free_y", facet_ncol = 3,
             palette = PAL, xlab = "library", ylab = "% of reads",
             # three facets share the width, so ten labels per panel need angling
             theme_args = list(axis.text.x = element_text(angle = 45, hjust = 1)),
             title = "Alignment quality falls off only at the two lowest input rungs")
save_fig(p, "R01_alignment", width = 11, height = 3.4)
```

Mechanically the run is sound: quality scores clean, no reads flagged poor quality,
strandedness auto-inferred `unstranded` on all ten (correct for FLASH-seq),
mitochondrial 1.1–3.2 %, intergenic 2.8–3.9 %, 5′/3′ bias 1.14–1.31.

The two 60 pg libraries and the 30 pg pair carry visibly more "unmapped, too short"
reads. Sections 5 and 6 show this has **two separate causes** that happen to land on
neighbouring rungs — a contaminant confined to two wells, and poly-G that scales
with input.

## 2. Sensitivity — genes detected across the titration

The obvious worry with a gene-detection drop is that it is really a depth
difference. It is not: `03_gene_detection.py` also reports the expected detection
count after binomial thinning to the shallowest library's total, and the split
survives it.

```{r det-table}
det <- read.delim(file.path(RES, "gene_detection.tsv"))
det <- det[match(ORDER, det$library), ]
det$short        <- shorten(det$library)
det$input_amount <- as_level(det$input_amount)
det[, c("library", "input_amount", "replicate", "total_counts",
        "genes_ge1", "genes_rarefied", "genes_tpm_gt1")]
```

```{r fig2, fig.width=11, fig.height=3.8}
p_left <- ScatterPlot(det, x = "input_pg", y = "genes_ge1",
                      color_by = "input_amount",
                      palette = PAL, xlab = "input (pg, log scale)",
                      ylab = "genes with >=1 count",
                      title = "detection vs input") +
  scale_x_log10() +
  geom_point(aes(x = input_pg, y = genes_rarefied), data = det,
             shape = 21, fill = NA, colour = "grey30", size = 2.6, stroke = 0.9)

fig2_dat <- rbind(
  data.frame(short = det$short, kind = "as sequenced", genes = det$genes_ge1),
  data.frame(short = det$short, kind = "rarefied",     genes = det$genes_rarefied))

p_right <- bar_lib(fig2_dat, x = "short", y = "genes", group_by = "kind",
                   palette = "Set2", xlab = "library",
                   ylab = "genes with >=1 count",
                   title = "depth is not the explanation")

p <- patchwork::wrap_plots(p_left, p_right, widths = c(1, 1.25))
save_fig(p, "R02_sensitivity", width = 11, height = 3.8)
p
```

Filled points are as sequenced, open circles are rarefied to the shallowest
library's total (16.5 M). 30 ng, 3 ng and 1.5 ng are **indistinguishable from one
another**; performance then collapses at pg scale.

**The design's blind spot:** there is no rung between 1.5 ng and 60 pg — a 25-fold
gap — and that gap is exactly where the method stops working.

## 3. Replicate concordance — where the titration actually breaks

```{r conc-table}
conc <- read.delim(file.path(RES, "replicate_concordance.tsv"))
# Mark any rung whose pair includes an excluded library, so its r cannot be read
# off this table as if it meant something.
conc$verdict <- ifelse(conc$rep1 %in% EXCLUDED | conc$rep2 %in% EXCLUDED,
                       "EXCLUDED PAIR", "ok")
conc
```

**The 60 pg row is not usable.** A8 was excluded (section 7), so that *r* is a
correlation against an excluded library, and the rung has no other replicate. The
conclusion does not rest on it: the **clean** 30 pg pair shows the same collapse
independently, *r* = 0.8435 against 0.980–0.987 for the ng rungs.

```{r tpm}
tpm <- read.delim(file.path(ROOT, "data/flashseq/results/star_rsem/rsem.merged.gene_tpm.tsv"),
                  check.names = FALSE)
rownames(tpm) <- tpm$gene_id
tpm <- tpm[, ORDER, drop = FALSE]          # drop gene_id / transcript_id(s)
# Same transform the Python side used for its scatter: log2(TPM + 1), over genes
# expressed anywhere. Genes silent in every library carry no information about
# concordance and would inflate every correlation towards 1.
expressed <- rowSums(tpm) > 0
lt <- log2(as.matrix(tpm[expressed, ]) + 1)
cat(nrow(lt), "genes expressed in at least one library\n")
```

```{r fig3, fig.width=12, fig.height=3.2}
# One panel per rung: replicate 1 against replicate 2, over genes expressed in
# both -- the same set the r in the table above is computed over.
fig3_dat <- do.call(rbind, lapply(seq_len(nrow(conc)), function(i) {
  a <- conc$rep1[i]; b <- conc$rep2[i]
  both <- lt[, a] > 0 & lt[, b] > 0
  data.frame(rep1 = lt[both, a], rep2 = lt[both, b],
             rung = sprintf("%s  (r = %.3f)%s", conc$input_amount[i],
                            conc$pearson_log2tpm[i],
                            if (conc$verdict[i] != "ok") "  [EXCLUDED PAIR]" else ""))
}))
fig3_dat$rung <- factor(fig3_dat$rung, levels = unique(fig3_dat$rung))

p <- CorPlot(fig3_dat, x = "rep1", y = "rep2", split_by = "rung",
             pt_size = 0.25, alpha = 0.25, add_smooth = TRUE,
             anno_items = c("eq", "r2"),
             xlab = "replicate 1, log2(TPM+1)", ylab = "replicate 2, log2(TPM+1)",
             ncol = 5)
save_fig(p, "R03_replicates", width = 12, height = 3.2)
p
```

The three ng rungs are a tight diagonal; the two pg rungs fan out. **The 60 pg panel
is labelled `[EXCLUDED PAIR]`** — it is a correlation against A8, which is excluded,
and there is no replacement replicate. The clean 30 pg panel is what carries the
conclusion.

## 4. rRNA — the pipeline's number is wrong, and wrong in a knowable direction

nf-core reports rRNA as the read fraction over genes annotated `gene_biotype "rRNA"`
in the Ensembl GTF. On GRCm39 that annotation holds 354 genes which are **essentially
all 5S** (`n-R5s*`) plus one `Rn18s-rs5` relic — there is **no `Rn45s`, `Rn28s` or
`Rn5-8s` gene at all**, because the rDNA array is collapsed out of the primary
assembly. So the reported 0.7–1.4 % measures 5S, not rRNA.

`01_rrna_kmer_screen.py` re-measures against `unique_rRNA_mouse.v2.fa`, which carries
the NCBI 47S pre-rRNA unit `BK000964.3:1-13403`.

```{r rrna-table}
rr <- read.delim(file.path(RES, "rrna_kmer.tsv"))
rr <- rr[match(ORDER, rr$library), ]
rr$short <- shorten(rr$library)
rr$nfcore_biotype_rRNA_pct <- qc$nfcore_biotype_rRNA_pct[match(rr$library, qc$library)]
rr$input_amount <- as_level(qc$input_amount[match(rr$library, qc$library)])
rr[, c("library", "input_amount", "reads_screened", "nfcore_biotype_rRNA_pct",
       "pct_rRNA_47S", "pct_rRNA_ensembl_only", "pct_rRNA_total")]
```

```{r fig4, fig.width=8, fig.height=3.6}
fig4_dat <- rbind(
  data.frame(short = rr$short, source = "nf-core: Ensembl gene_biotype rRNA",
             pct = rr$nfcore_biotype_rRNA_pct),
  data.frame(short = rr$short, source = "k-mer: Ensembl records only",
             pct = rr$pct_rRNA_ensembl_only),
  data.frame(short = rr$short, source = "k-mer: 47S unit (BK000964.3)",
             pct = rr$pct_rRNA_47S))
fig4_dat$source <- factor(fig4_dat$source, levels = c(
  "nf-core: Ensembl gene_biotype rRNA", "k-mer: Ensembl records only",
  "k-mer: 47S unit (BK000964.3)"))

p <- bar_lib(fig4_dat, x = "short", y = "pct", group_by = "source",
             palette = "Dark2", xlab = "library", ylab = "% of reads",
             title = "What the annotation-based rRNA number structurally cannot see")
save_fig(p, "R04_rrna", width = 8, height = 3.6)
```

```{r rrna-share}
share <- 100 * sum(rr$pct_rRNA_47S) / sum(rr$pct_rRNA_total)
ratio <- sum(rr$pct_rRNA_total) / sum(rr$nfcore_biotype_rRNA_pct)
cat(sprintf("%.1f%% of the rRNA signal comes from the 47S unit Ensembl does not annotate.\n", share))
cat(sprintf("The honest number is %.1fx the one nf-core reports.\n", ratio))
```

These are exact 31-mer matches sampled every 10 bases with no mismatch tolerance,
so they are a **lower bound**. Section 4b aligns the same reads instead.

## 4b. The same measurement as VASA, not merely a similar one

The k-mer screen is a lower bound and the VASA figure it needs to be compared with
(21.39 % over `ZHA9292A1`) came from `bwa`. Two different methods is not a
comparison. `05_rrna_bwa.sh` therefore runs **the VASA pipeline's own rRNA stage** —
`a_Mapping/ribo-bwamem.sh` and `riboread-selection.py`, unmodified, `bwa aln` **and**
`bwa mem`, the same `unique_rRNA_mouse.v2.fa` — over the FLASH-seq reads.

Three choices in that script are not obvious, and all three are checked rather than
assumed:

* **`stranded = n`, not `y`.** The flag decides whether a reverse-mapping read may be
  called ribosomal. VASA is stranded, so `y` is right *there*; FLASH-seq is not, so
  `y` would discard half the rRNA for no reason. The same BAM is re-selected with `y`
  and reported alongside.
* **Two arms, raw and trimmed**, because VASA's 21.39 % is over trimmed reads.
* **R1 only**, because VASA has one biological read per fragment and letting
  FLASH-seq have two would bias in its favour.

```{r bwa-table}
rb <- read.delim(file.path(RES, "rrna_bwa.tsv"), check.names = FALSE)
tr <- rb[rb$arm == "trimmed", ]
tr <- tr[match(ORDER, tr$library), ]
tr$short        <- shorten(tr$library)
tr$input_amount <- as_level(tr$input_amount)

# From data/PM26037/out/logs/step3_report.txt, job 50788552 -- the ALL rows of its
# two tables. Same scripts, same reference, so these are directly comparable.
VASA_RIBO <- 21.39
VASA_COMP <- c(`5ETS` = 17.1, `18S` = 5.9, ITS1 = 4.6, `5.8S` = 1.9,
               ITS2 = 7.6, `28S` = 54.4, `3ETS` = 0.9)
SUB <- names(VASA_COMP)

tr[, c("library", "input_amount", "reads_in", "read_len_median", "ribo_pct",
       "ribo_pct_stranded_y", "kmer_pct", "aln_only", "mem_only", "both", "fwd_pct")]
```

```{r fig8, fig.width=11.5, fig.height=3.8}
fig8a_dat <- rbind(
  data.frame(short = tr$short, method = "k-mer (exact match, lower bound)", pct = tr$kmer_pct),
  data.frame(short = tr$short, method = "bwa aln + mem (VASA's own stage)", pct = tr$ribo_pct))
fig8a_dat$method <- factor(fig8a_dat$method, levels = unique(fig8a_dat$method))

pA <- bar_lib(fig8a_dat, x = "short", y = "pct", group_by = "method",
              palette = "Set1", xlab = "library", ylab = "% of reads",
              title = "rRNA content, trimmed reads") +
  geom_hline(yintercept = VASA_RIBO, linetype = "dashed",
             colour = "#1f4e79", linewidth = 0.7) +
  annotate("text", x = Inf, y = VASA_RIBO, hjust = 1.05, vjust = -0.5,
           label = sprintf("VASA %.1f%%", VASA_RIBO), colour = "#1f4e79", size = 3)

# Composition as a share of ALL reads. Comparing composition percentages alone
# misleads when the totals differ 4x: a subunit can be a bigger slice of a much
# smaller cake.
fig8b_dat <- rbind(
  data.frame(subunit = SUB, method = "FLASH-seq (mean of 10)",
             pct = sapply(SUB, function(s) mean(tr[[paste0("abs_", s)]]))),
  data.frame(subunit = SUB, method = "VASA-seq ZHA9292A1",
             pct = as.numeric(VASA_RIBO * VASA_COMP / 100)))
fig8b_dat$subunit <- factor(fig8b_dat$subunit, levels = SUB)

pB <- bar_lib(fig8b_dat, x = "subunit", y = "pct", group_by = "method",
              palette = "Set1", xlab = "47S subunit", ylab = "% of all reads",
              title = "Where the difference actually sits")

p <- patchwork::wrap_plots(pA, pB, widths = c(1.15, 1))
save_fig(p, "R08_rrna_bwa", width = 11.5, height = 3.8)
p
```

```{r bwa-summary}
cat(sprintf("FLASH-seq trimmed: %.2f-%.2f%% ribosomal\n", min(tr$ribo_pct), max(tr$ribo_pct)))
cat(sprintf("VASA-seq:          %.2f%% whole-library (18.0-25.1%% across its 12 real cells)\n", VASA_RIBO))
cat(sprintf("  -> VASA carries %.1fx the rRNA.\n\n", VASA_RIBO / mean(tr$ribo_pct)))
cat(sprintf("mature 28S: %.2f%% of VASA reads vs %.2f%% of FLASH-seq (%.1fx)\n",
            VASA_RIBO * VASA_COMP[["28S"]] / 100, mean(tr$abs_28S),
            (VASA_RIBO * VASA_COMP[["28S"]] / 100) / mean(tr$abs_28S)))
cat(sprintf("5'ETS:      %.2f%% vs %.2f%% (%.1fx)\n",
            VASA_RIBO * VASA_COMP[["5ETS"]] / 100, mean(tr$abs_5ETS),
            (VASA_RIBO * VASA_COMP[["5ETS"]] / 100) / mean(tr$abs_5ETS)))
cat("The gap is in MATURE rRNA, not pre-rRNA -- which is what a poly-A-primed\n")
cat("protocol versus a total-RNA protocol predicts.\n\n")
cat(sprintf("stranded=y would have reported %.2f%% instead of %.2f%%.\n",
            mean(tr$ribo_pct_stranded_y), mean(tr$ribo_pct)))
cat(sprintf("forward-strand share of ribosomal reads: %.1f-%.1f%% -- i.e. unstranded,\n",
            min(tr$fwd_pct), max(tr$fwd_pct)))
cat("confirming n is the right flag.\n\n")
cat(sprintf("aln-only detection: %s reads of %s. At 151 nt bwa aln adds almost nothing --\n",
            format(sum(tr$aln_only), big.mark = ","),
            format(sum(tr$aln_only + tr$mem_only + tr$both), big.mark = ",")))
cat("the mirror image of VASA, where reads are short and aln-only was ~49% of detection.\n")
```

The 5'ETS share is high (32–47 % of ribosomal reads), and on the VASA side a high
5'ETS share once turned out to be a poly-T artefact. It is **not** one here: the
`ets5_peak_bin_pct` column shows the 5'ETS hits spread across the whole 4 kb,
busiest 200 nt window 12.3–16.0 %, against 88.9 % in the case that *was* an artefact.

```{r ets-check, fig.width=7, fig.height=3.2}
p <- bar_lib(tr, x = "short", y = "ets5_peak_bin_pct", fill_by = "input_amount",
             palette = PAL, xlab = "library",
             ylab = "% of 5'ETS hits in busiest 200 nt window",
             title = "5'ETS signal is spread, not piled up (artefact case was 88.9%)") +
  geom_hline(yintercept = 88.9, linetype = "dotted", colour = "firebrick")
save_fig(p, "R08b_ets_check", width = 7, height = 3.2)
```

## 5. Contamination — human CALB1, and it is a well effect

FastQC flagged a sequence making up 15.9 % of A8 and could only say "No Hit".
`02_contaminant_check.py` resolves every flagged sequence against mouse GRCm39, the
rRNA reference, ERCC92 and human GRCh38, then **re-counts each one in every
library** — the second step matters because FastQC only reports a sequence above
~0.1 % of a file, so it cannot distinguish "absent" from "below the floor".

```{r contam}
ov <- read.delim(file.path(RES, "overrepresented.tsv"))
hum <- ov[startsWith(as.character(ov$source), "human_GRCh38"), ]
# A fragment's R1 and R2 are separate rows and pct_in_library is the larger of the
# two, so summing THAT column double-counts every pair. Sum pct_R1.
hum_tot <- tapply(hum$pct_R1, hum$library, sum)
fig5_dat <- data.frame(
  short        = shorten(ORDER),
  input_amount = as_level(meta$input_amount),
  well         = meta$well,
  pct_human    = as.numeric(hum_tot[ORDER]))
fig5_dat$pct_human[is.na(fig5_dat$pct_human)] <- 0
fig5_dat
```

```{r fig5, fig.width=8, fig.height=3.6}
p <- bar_lib(fig5_dat, x = "short", y = "pct_human", fill_by = "input_amount",
             palette = PAL, xlab = "library", ylab = "% of R1 reads, human",
             label = "well", label_size = 2.8,
             title = "Human CALB1 tracks adjacent wells (G:1, H:1), not input amount")
save_fig(p, "R05_contamination", width = 8, height = 3.6)
```

The 30 pg pair (wells A:2/B:2) is *cleaner* than the 30 ng pair, so this is **not**
"low input amplifies background". A7 and A8 are wells **G:1 and H:1 — adjacent**.
This is localised contamination introduced during prep. The STP's own `fastq_screen`,
which never saw this pipeline, agrees: A8 is 79.9 % mouse / 42.4 % human against
~98 % / ~26 % for A1–A6.

## 6. Artefacts — adapter read-through and poly-G

```{r artefacts}
probe <- function(src, col = "pct_R1") {
  v <- ov[[col]][ov$source == src]
  l <- ov$library[ov$source == src]
  as.numeric(tapply(v, l, max)[ORDER])
}
fig6_dat <- rbind(
  data.frame(short = shorten(ORDER), artefact = "adapter read-through (R1)",
             pct = probe("probe_adapter_P7_readthrough")),
  data.frame(short = shorten(ORDER), artefact = "poly-G, any 30 nt run (R1)",
             pct = probe("probe_polyG_30")),
  data.frame(short = shorten(ORDER), artefact = "poly-G, any 30 nt run (R2)",
             pct = probe("probe_polyG_30", "pct_R2")))
```

```{r fig6, fig.width=9, fig.height=3.6}
p <- bar_lib(fig6_dat, x = "short", y = "pct", group_by = "artefact",
             palette = "Dark2", xlab = "library", ylab = "% of reads",
             title = "Both artefacts rise as input falls")
save_fig(p, "R06_artefacts", width = 9, height = 3.6)
```

Read-through is strictly R1 and rises from 16–20 % at 30 ng to ~32 % at 30 pg —
shorter inserts from less material. Poly-G is the NovaSeq X two-colour dark-cycle
artefact: with no signal, the base caller emits G.

### What trimming actually removes

```{r fig7, fig.width=7.5, fig.height=3.4}
te <- read.delim(file.path(RES, "trim_effect.tsv"))
agg <- aggregate(cbind(pct_raw, pct_trimmed) ~ kind, data = te, FUN = mean)
fig7_dat <- rbind(
  data.frame(kind = agg$kind, pass = "raw",     pct = agg$pct_raw),
  data.frame(kind = agg$kind, pass = "trimmed", pct = agg$pct_trimmed))
fig7_dat$pass <- factor(fig7_dat$pass, levels = c("raw", "trimmed"))

p <- bar_lib(fig7_dat, x = "kind", y = "pct", group_by = "pass",
             palette = "Paired", xlab = "sequence class",
             ylab = "mean % per read file",
             title = "Comparing FastQC to FastQC: nothing much is removed")
save_fig(p, "R07_trim_effect", width = 7.5, height = 3.4)
```

None of the three classes is meaningfully removed. `06_trim_options.sh` then tested
the two fixes that were proposed for this, and **withdrew one of them**:

* `--nextseq-trim=20` / `--2colour 20` removes **2.7 %** of A10's read pairs.
* But the "read-through that survives trimming" and the poly-G are **the same
  reads** — only 1.14 % of A10's R1 begins mid-mosaic-end, and **1,019 of those
  1,032 have a poly-G R2 mate**. One no-insert artefact seen from two ends.
* Adding a second adapter pattern `CTTATACACATCT` is **harmful**: at 13 nt it beats
  `CTGTCTCTTATA`'s 12 nt in cutadapt's best-match contest on ordinary read-through,
  cuts 7 nt later, and leaves mosaic-end sequence on **37.18 %** of R1 reads instead
  of 0.10 %.

So the gain from better trimming is a cleaner denominator, not recovered data —
those pairs have no insert to recover.

## 7. What this means for the VASA-seq comparison

VASA-seq operates at single-cell input, ~10–30 pg. The comparable FLASH-seq rungs
are the two lowest, and they are also the two most compromised.

**A8 was excluded on 2026-07-27.** The decision lives in `sample_metadata.tsv` as
`qc_verdict`, so it travels with the data rather than living only in prose.

| rung | libraries | verdict |
|---|---|---|
| 30 pg | A9, A10 | **`ok`** — the cleanest low-input pair |
| 60 pg | A7 | **`caveat`** — 3.6 % human CALB1, filterable |
| 60 pg | A8 | **`exclude`** — 18.3 % of the library is not mouse |
| ≥1.5 ng | A1–A6 | **`ok`**, but not input-comparable to VASA; the ceiling reference |

Excluding A8 costs nothing in the headline conclusion — the cliff is shown
independently by the clean 30 pg pair — but it leaves the 60 pg rung with **no
replicate at all**, which is why section 3's 60 pg row must not be quoted.

**Quote FLASH-seq's rRNA fraction from section 4b**, never the nf-core biotype
number. On the 30 pg rung that is **A9 5.26 % / A10 4.02 %** against VASA's 21.39 %.

## 8. Caveats about the run's saved outputs

```{r caveats}
gtf <- file.path(ROOT, "data/flashseq/results/genome/Mus_musculus.GRCm39.116.filtered.gtf")
bed <- file.path(ROOT, "data/flashseq/results/genome/Mus_musculus.GRCm39.116.filtered.bed")
sz  <- file.info(gtf)$size
cat(sprintf("filtered.gtf: %s bytes, %s a multiple of 64 KiB\n",
            format(sz, big.mark = ","), if (sz %% 65536 == 0) "EXACTLY" else "not"))
cat(sprintf("filtered.bed rows: %s\n",
            format(length(readLines(bed)), big.mark = ",")))
stopifnot(sz %% 65536 == 0)   # the signature of the partial copy; see below
```

**The saved GTF is truncated to chromosome 1** — 4,742 gene ids, and a size that is
an exact multiple of 64 KiB, the signature of a partial copy. The sibling
`.filtered.bed` and the RSEM reference are complete, so **this run's quantification
is unaffected**, but that GTF would silently produce a chr1-only analysis if reused.
The assertion above exists so this cannot be forgotten.

**The STAR index was not saved** despite `--save_reference`. The work directory has
been deleted, so a re-run rebuilds it (22 min, 65 GB peak).

## 9. Where `scplotter` proper applies

Everything above is a per-library summary, which is why it is drawn with `plotthis`.
Here the ten libraries genuinely can be treated as ten samples in a Seurat object,
and `scplotter::CellDimPlot` is doing what it was built for: a PCA of the libraries
in expression space, coloured by input amount.

This panel has **no Python counterpart** — it is new, and it earns its place by
putting the A8 decision on an independent footing. The exclusion was argued from
targeted sequence counting (`02_contaminant_check.py` searching for one human
fragment). An unsupervised ordination of whole transcriptomes knows nothing about
CALB1, so if it separates A8 for the same reason, that is a second, independent
witness.

It does. Read the figure this way:

* **A1–A6 form one tight cluster** — the six ng libraries, consistent with their
  *r* = 0.980–0.987.
* **A8 is a large outlier on PC1**, far from everything including its own 60 pg
  partner.
* **A7 sits with the ng cluster**, not with A8 — same input amount, same rung,
  adjacent well, completely different position.

That last pair is the whole argument: if 60 pg input were what made A8 strange, A7
would be strange with it. It is not. The separation tracks the contamination, not
the rung — which is what the well pattern (G:1 / H:1) said, arrived at here by a
method that was never told about it. **A9 and A10 are also displaced**, along PC2
rather than PC1, and that *is* the input effect: they are the genuinely
RNA-limited pair.

```{r pca}
# Genes expressed above 1 TPM in at least half the libraries, so the PCA is not
# driven by presence/absence in the shallow ones.
keep <- rowSums(tpm > 1) >= 5
lmat <- log2(as.matrix(tpm[keep, ]) + 1)
pca  <- prcomp(t(lmat), scale. = TRUE)
ve   <- 100 * pca$sdev^2 / sum(pca$sdev^2)

pcs <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2],
                  library = rownames(pca$x))
pcs$short        <- shorten(pcs$library)
pcs$input_amount <- as_level(meta$input_amount[match(pcs$library, meta$library)])
pcs$qc_verdict   <- meta$qc_verdict[match(pcs$library, meta$library)]
cat(sprintf("%d genes; PC1 %.1f%% of variance, PC2 %.1f%%\n", sum(keep), ve[1], ve[2]))
```

```{r fig9, fig.width=6.8, fig.height=4.8}
have_scplotter <- requireNamespace("scplotter", quietly = TRUE) &&
                  requireNamespace("Seurat", quietly = TRUE)

if (have_scplotter) {
  # The ten libraries as ten "cells" -- the one place in this document where
  # scplotter's own API is the right tool rather than a contrivance.
  suppressPackageStartupMessages({library(Seurat); library(scplotter)})
  obj <- CreateSeuratObject(counts = as.matrix(tpm[keep, ]), assay = "RNA")
  obj$input_amount <- pcs$input_amount[match(colnames(obj), pcs$library)]
  obj$qc_verdict   <- pcs$qc_verdict[match(colnames(obj), pcs$library)]
  obj[["pca"]] <- CreateDimReducObject(
    embeddings = pca$x[colnames(obj), 1:2, drop = FALSE], key = "PC_", assay = "RNA")
  p <- CellDimPlot(obj, reduction = "pca", group_by = "input_amount",
                   palette = PAL, pt_size = 4, theme = "theme_this",
                   title = "PCA of the ten libraries (scplotter::CellDimPlot)")
} else {
  # Same coordinates, same colouring, drawn by scplotter's engine instead. The
  # figure is the figure; only the function differs.
  p <- plotthis::DimPlot(pcs, dims = c("PC1", "PC2"), group_by = "input_amount",
                         palette = PAL, pt_size = 4,
                         title = "PCA of the ten libraries (plotthis::DimPlot)",
                         xlab = sprintf("PC1 (%.1f%%)", ve[1]),
                         ylab = sprintf("PC2 (%.1f%%)", ve[2]))
}
p <- p + ggrepel::geom_text_repel(
  data = pcs, aes(x = PC1, y = PC2,
                  label = ifelse(qc_verdict == "ok", as.character(short),
                                 paste0(short, " [", qc_verdict, "]"))),
  size = 3, seed = 1, min.segment.length = 0.2, max.overlaps = Inf,
  inherit.aes = FALSE)
save_fig(p, "R09_library_pca", width = 6.8, height = 4.8)
p
```

```{r pca-note, echo=FALSE, results="asis"}
cat(if (have_scplotter)
  "Drawn with `scplotter::CellDimPlot`.\n" else
  paste0("**`scplotter` is not installed here, so this panel fell back to ",
         "`plotthis::DimPlot` — the same coordinates and the same engine.** ",
         "The blocker is documented in `install_r_deps.sh`: scplotter `Imports` ",
         "`scRepertoire`, a TCR/BCR repertoire package this analysis has no use ",
         "for, and installing it fails twice over — R's conda `libcurl` cannot ",
         "reach Bioconductor's archive hosts (system `curl` can), and the `gsl` ",
         "R package needs a C library that is not installed system-wide. ",
         "Nothing else in this document depends on it.\n"))
```

## Session

```{r session}
sessionInfo()
```
